# Additonal plotting tools


```{r, include=FALSE, cache=F, echo=F}
library(data.table)
library(gridExtra)
library(ggplot2)
library(magrittr)


# data
ind <- fread('extdata/CPI_HDI.csv')

mysize <- 15
mytheme <- theme(
    axis.title = element_text(size=mysize), 
    axis.text = element_text(size=mysize),
    legend.title = element_text(size=mysize),
    legend.text = element_text(size=mysize)
    ) + theme_bw()

```

## Plotting themes

Themes control non-data parts of your plots, such as:

* Overall appearance
* Axes
* Plot title
* Legends

They control the appearance of your plots (size, color, position) but not how the data is represented.
One example is `theme_bw` which provides a white background.

```{r,  fig.height=5, fig.width=6}
ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_point() +
  theme_bw()
```

Other complete themes are: `theme_classic`, `theme_minimal`, `theme_light`, `theme_dark`.

More themes are available in the `ggthemes` package.
```{r, fig.height=4.5, fig.width=5.5, fig.cap = 'Example of Wall Street Journal theme'}
library(ggthemes)
ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_point() +
  theme_wsj() + scale_colour_wsj("colors6", "")
```


## Axes

The appearance of the axis titles can be controlled with the global variable `axis.title`,
or independently for $x$ and $y$ using `axis.title.x` and `axis.title.y`.

```{r,  fig.height=4.5, fig.width=6}
ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_jitter() +
  theme(axis.title = element_text(size=20, color = 'red'))
```

Similar for `axis.text`:
```{r, fig.height=4.5, fig.width=6}
ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_jitter() +
  theme(axis.text = element_text(size=20, color = 'blue'))
```


### Axis elements

The axis elements control the appearance of the axes:

Element             | Setter            | Description
--------------------|-------------------|---------------------------
axis.line           | `element_line()`  | line parallel to axis (hidden in default themes)
axis.text           | `element_text()`  | tick labels
axis.text.x         | `element_text()`  | x-axis tick labels
axis.text.y         | `element_text()`  | y-axis tick labels
axis.title          | `element_text()`  | axis titles
axis.title.x        | `element_text()`  | x-axis title
axis.title.y        | `element_text()`  | y-axis title
axis.ticks          | `element_line()`  | axis tick marks
axis.ticks.length   | `unit()`          | length of tick marks

## Plot title

The appearance of the plot's title can be controlled with the variable `plot.title`.
```{r, fig.height=5, fig.width=6}
ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_jitter() +
  ggtitle('CPI vs HDI') +
  theme(plot.title = element_text(size=20, face = 'bold'))
```


## Legend

The appearance of the legend can be controlled with `legend.text` and `legend.title`.
```{r, fig.height=5, fig.width=7}
base <- ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_jitter()
base + theme(
  legend.text = element_text(size = 15),
  legend.title = element_text(size = 15, face = "bold")
)
```

The legend elements control the appearance of all legends. You can also modify the appearance of individual legends by modifying the same elements in `guide_legend()` or `guide_colourbar()`.

Element             | Setter                    | Description                                 |
--------------------|---------------------------|---------------------------------------------|
legend.background   |  `element_rect()`         | legend background                           |
legend.key          |  `element_rect()`         | background of legend keys                   |
legend.key.size     |  `unit()`                 | legend key size                             |
legend.key.height   |  `unit()`                 | legend key height                           |
legend.key.width    |  `unit()`                 | legend key width                            |
legend.margin       |  `unit()`                 | legend margin                               |
legend.text         |  `element_text()`         | legend labels                               |
legend.text.align   |  0--1                     | legend label alignment (0 = right, 1 = left)|
legend.title        |  `element_text()`         | legend name                                 |
legend.title.align  |  0--1                     | legend name alignment (0 = right, 1 = left) |



## Interactive plots

Interactive plots can improve presentations. The library `plotly` allows to interactively construct plots. We can simply construct a plot with `ggplot2` and pass this object to the function `ggplotly()`:

```{r,  eval=F, echo=T}
library(plotly)
p <- ggplot(mtcars, aes(factor(cyl), mpg)) +
        geom_boxplot()
ggplotly(p)
```

We can also save the generated `plotly` object into an `html` file as follows:

```{r,  echo=T, eval=F}
plotly_path <- "assets/fig/my_plotly_fig.html"
save_plotly_as_widget(ggplotly(p), file.path(getwd(), plotly_path))
```


